Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the `eechidna` `R` package.

### 2019-03-25 01-intro.Rmd
added data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, fig.pos = 'h', fig.align = 'center')
library(tidyverse)
library(eechidna)
library(ggplot2)
library(ggthemes)
library(knitr)
```

>>>>

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = F, fig.pos = 'h', fig.align = 'center')
library(tidyverse)
library(eechidna)
library(ggplot2)
library(ggthemes)
library(knitr)
library(nlme)
library(rgeos)
library(spdep)

# Data
data(tpp01)
data(tpp04)
data(tpp07)
data(tpp10)
data(tpp13)
data(tpp16)
data(abs2001)
data(abs2004)
data(abs2007)
data(abs2010)
data(abs2013)
data(abs2016)
```

```{r wrangle, include = F, message = F}
# Function to standardize variables
standardise_vars <- function(df) {
  hold <- df
  num_cols <- sapply(df, class) == 'numeric'
  df[, num_cols] <- lapply(df[, num_cols], scale)
  names(df) <- names(hold)
  df$LNP_Percent <- hold$LNP_Percent
  return(df)
}

# Combine and standardize
my_df <- bind_rows(
  left_join(tpp01, standardise_vars(abs2001) %>% dplyr::select(-c(UniqueID, Area, ends_with("NS"), Population)), by = c("DivisionNm", "StateAb"="State")) %>% mutate(year = "2001"),
  left_join(tpp04, standardise_vars(abs2004) %>% dplyr::select(-UniqueID), by = c("DivisionNm")) %>% mutate(year = "2004"),
  left_join(tpp07, standardise_vars(abs2007) %>% dplyr::select(-UniqueID), by = c("DivisionNm")) %>% mutate(year = "2007"),
  left_join(tpp10, standardise_vars(abs2010) %>% dplyr::select(-UniqueID), by = c("DivisionNm")) %>% mutate(year = "2010"),
  left_join(tpp13, standardise_vars(abs2013) %>% dplyr::select(-UniqueID), by = c("DivisionNm")) %>% mutate(year = "2013"),
  left_join(tpp16, standardise_vars(abs2016) %>% dplyr::select(-c(UniqueID, Area, ends_with("NS"), Population)), by = c("DivisionNm", "StateAb"="State")) %>% mutate(year = "2016")
) %>% 
  mutate(year = factor(year)) %>% 
  dplyr::select(-c(starts_with("Age"), StateAb, LNP_Votes, ALP_Votes, ALP_Percent, TotalVotes,
    Swing, InternetUse, InternetAccess, EnglishOnly, Other_NonChrist, OtherChrist, Volunteer, EmuneratedElsewhere, UniqueID, Catholic, Anglican))

# Create final df for modelling

factors_df <- my_df %>% 
  mutate(Education = BachelorAbv + HighSchool + Professional + Finance - Laborer - Tradesperson - DipCert,
    FamHouseSize = FamilyRatio + AverageHouseholdSize + Couple_WChild_House - Couple_NoChild_House -
      SP_House,
    PropertyOwned = Owned + Mortgage - Renting - PublicHousing,
    RentLoanPrice = MedianRent + MedianLoanPay,
    Incomes = MedianFamilyIncome + MedianHouseholdIncome + MedianPersonalIncome,
    Unemployment = Unemployed - LFParticipation) %>% 
  dplyr::select(-c(BachelorAbv, HighSchool, Professional, Finance, Laborer, Tradesperson, DipCert, FamilyRatio,
    AverageHouseholdSize, Couple_WChild_House, Couple_NoChild_House, SP_House, Owned, Mortgage, Renting,
    PublicHousing, MedianFamilyIncome, MedianHouseholdIncome, MedianPersonalIncome, MedianRent, 
    MedianLoanPay, Unemployed, LFParticipation))

# Now standardize factors

small_df <- bind_rows(
  factors_df %>% filter(year == "2001") %>% standardise_vars(),
  factors_df %>% filter(year == "2004") %>% standardise_vars(),
  factors_df %>% filter(year == "2007") %>% standardise_vars(),
  factors_df %>% filter(year == "2010") %>% standardise_vars(),
  factors_df %>% filter(year == "2013") %>% standardise_vars(),
  factors_df %>% filter(year == "2016") %>% standardise_vars()
)

# Order electorates in alphabetical order to match spatial matrix
model_df <- small_df %>% 
  arrange(year, DivisionNm)
```

```{r spatialweights, eval = F}
# Compute spatial weights matrix
sp_weights_matrix <- function(sF) {

dist_matrix <- function(shapefile) {
  dist_mat <- matrix(NA, nrow = 150, ncol = 150)
  rownames(dist_mat) <- sort(shapefile$elect_div)
  colnames(dist_mat) <- sort(shapefile$elect_div)

  for (i in 1:(nrow(dist_mat)-1)) {
    rname = rownames(dist_mat)[i]
    row_poly = shapefile %>% subset(elect_div == rname)

    for (j in (i+1):ncol(dist_mat)) {

      cname = rownames(dist_mat)[j]
      col_poly = shapefile %>% subset(elect_div == cname)
      dist = gDistance(row_poly, col_poly)
      dist_mat[i,j] = dist

    }
    print(i)
  }

  # Now copy to lower triange
  for (i in 2:nrow(dist_mat)) {
    for (j in 1:(i-1)) {
      dist_mat[i,j] = dist_mat[j,i]
    }
  }

  # Check it is symmetric
  if(!isSymmetric(dist_mat)) {
    print("Warning! Matrix is not symmetric. Error has occured.")
  }

  return(dist_mat)
}

my_dist_mat <- dist_matrix(sF)

# S matrix

s_mat <- function(dist_mat) {
  s_mat <- dist_mat

  for (i in 1:nrow(dist_mat)) {
    for (j in 1:nrow(dist_mat)) {
      a = dist_mat[i,j]

      if (is.na(a)) {
        b = 0
      } else {
        b = ifelse(a == 0, 1, 0)
      }

      s_mat[i,j] = b

    }
  }

  return(s_mat)
}

my_smat <- s_mat(my_dist_mat)

# Turn into W matrix

w_mat <- function(s_mat) {

  w_mat <- s_mat
  rsums = rowSums(s_mat)

  for (i in 1:nrow(s_mat)) {
    w_mat[i,] <- s_mat[i,]/rsums[i]
  }

  return(w_mat)
}

my_wmat <- w_mat(my_smat)

# Turn into listw

my_listw <- mat2listw(my_wmat)

return(my_listw)

}

# Get spatial weights
sF_16 <- sF_download(2016)
sF_13 <- sF_download(2013)
sF_10 <- sF_download(2010)
sF_07 <- sF_download(2007)
sF_04 <- sF_download(2004)
sF_01 <- sF_download(2001)

sp_weights_16 <- sp_weights_matrix(sF_16)
sp_weights_13 <- sp_weights_matrix(sF_13)
sp_weights_10 <- sp_weights_matrix(sF_10)
sp_weights_07 <- sp_weights_matrix(sF_07)
sp_weights_04 <- sp_weights_matrix(sF_04)
sp_weights_01 <- sp_weights_matrix(sF_01)
```

```{r FGLS, include = F}
# Instead of running spatial weights
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_01.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_04.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_07.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_10.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_13.rda")
load("/Users/Jeremy/Documents/R/eechidna-modelling/data/sp_weights_16.rda")

# Function for FGLS
my_fgls <- function(my_formula, my_data, sp_weights) {
  # Spatial weights matrix
  w_mat <- listw2mat(sp_weights)

  # Get OLS residuals
  ols_model <- lm(my_formula, my_data)
  my_res <- ols_model$residuals

  # Solve for rho
  res_model <- lm(my_res ~ w_mat%*%my_res)
  rho <- res_model$coefficients[2]
  rho_df <- data.frame(estimate = rho, se = summary(res_model)$coefficients[2,2], p = summary(res_model)$coefficients[2,4])

  # Transform data for GLS
  trans_mat <- diag(nrow(my_data)) - rho*w_mat
  gls_data <- data.frame(LNP_Percent = trans_mat %*% my_data$LNP_Percent,
    Intercept = trans_mat %*% rep(1,nrow(my_data))) %>% 
    bind_cols(as.data.frame(trans_mat %*% as.matrix(my_data %>% select(-LNP_Percent))))

  # GLS model
  my_formula <- formula(paste0(my_formula,  " - 1"))
  gls_model <- gls(my_formula, gls_data)
  gls_model$rho_df <- rho_df
  gls_model$gls_data <- gls_data
  gls_model$my_data <- my_data

  return(gls_model)
}

## Run full models for each year
full_formula = "LNP_Percent ~ ."

# 2016
glsmod16 <- my_fgls(full_formula, 
  my_data = model_df %>% filter(year == "2016") %>% select(-c(DivisionNm, year)),
  sp_weights = sp_weights_16)

# 2013
glsmod13 <- my_fgls(full_formula, 
  my_data = model_df %>% filter(year == "2013") %>% select(-c(DivisionNm, year)),
  sp_weights = sp_weights_13)

# 2010
glsmod10 <- my_fgls(full_formula, 
  my_data = model_df %>% filter(year == "2010") %>% select(-c(DivisionNm, year)),
  sp_weights = sp_weights_10)

# 2007
glsmod07 <- my_fgls(full_formula, 
  my_data = model_df %>% filter(year == "2007") %>% select(-c(DivisionNm, year)),
  sp_weights = sp_weights_07)

# 2004
glsmod04 <- my_fgls(full_formula, 
  my_data = model_df %>% filter(year == "2004") %>% select(-c(DivisionNm, year)),
  sp_weights = sp_weights_04)

# 2001
glsmod01 <- my_fgls(full_formula, 
  my_data = model_df %>% filter(year == "2001") %>% select(-c(DivisionNm, year)),
  sp_weights = sp_weights_01)

## Visualise coefficients and significance

coef_df <- bind_rows(
  data.frame(variable = glsmod16$coefficients %>% names, estimate = glsmod16$coefficients %>% unname, se = summary(glsmod16)$tTable[, "Std.Error"] %>% unname, p = summary(glsmod16)$tTable[, "p-value"] %>% unname, year = 2016),
  data.frame(variable = glsmod13$coefficients %>% names, estimate = glsmod13$coefficients %>% unname, se = summary(glsmod13)$tTable[, "Std.Error"] %>% unname, p = summary(glsmod13)$tTable[, "p-value"] %>% unname, year = 2013),
  data.frame(variable = glsmod10$coefficients %>% names, estimate = glsmod10$coefficients %>% unname, se = summary(glsmod10)$tTable[, "Std.Error"] %>% unname, p = summary(glsmod10)$tTable[, "p-value"] %>% unname, year = 2010),
  data.frame(variable = glsmod07$coefficients %>% names, estimate = glsmod07$coefficients %>% unname, se = summary(glsmod07)$tTable[, "Std.Error"] %>% unname, p = summary(glsmod07)$tTable[, "p-value"] %>% unname, year = 2007),
  data.frame(variable = glsmod04$coefficients %>% names, estimate = glsmod04$coefficients %>% unname, se = summary(glsmod04)$tTable[, "Std.Error"] %>% unname, p = summary(glsmod04)$tTable[, "p-value"] %>% unname, year = 2004),
  data.frame(variable = glsmod01$coefficients %>% names, estimate = glsmod01$coefficients %>% unname, se = summary(glsmod01)$tTable[, "Std.Error"] %>% unname, p = summary(glsmod01)$tTable[, "p-value"] %>% unname, year = 2001)
)
```
